Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-02-11 ms/cover_letter/plosHIVcover_resubm2.tex
stuff for resubmission

\revcomment{
The outstanding issue is that the paper remains quite challenging to
read. It is just not written as clearly as it could be. In many
paragraphs, the authors launch into discussion of complex points with
long, multi-clause sentences, instead of explaining their motivation
and results in simple, concise phrases. Too often the authors do not
make it clear WHY they did something. Reading the original papers by
Fraser’s group on this topic (PNAS 2007 and PLoS Comp Biol 2011)
provide good examples of clear writing.
}
\response{This is a very general comment, but we have gone
through the entire manuscript and tried to clarify it.}

@@

\response{We have added detail; we believe the presentation of our methods can stand on its own
(although we still relegate minor technical details to the supplementary materials, so as not
to overwhelm the reader). We have also changed presentation more generally so that we don't lead by
saying ``as in Shireff et al \ldots'', which might make readers think that they need to know Shirreff's
material, although we still try to be clear where we are following Shirref et al/Champredon et al and
where we are breaking new ground.}

>>>>

\revcomment{
The outstanding issue is that the paper remains quite challenging to
read. It is just not written as clearly as it could be. In many
paragraphs, the authors launch into discussion of complex points with
long, multi-clause sentences, instead of explaining their motivation
and results in simple, concise phrases. Too often the authors do not
make it clear WHY they did something. Reading the original papers by
Fraser’s group on this topic (PNAS 2007 and PLoS Comp Biol 2011)
provide good examples of clear writing.
}

\response{This is a very general comment, but we have gone
through the entire manuscript and tried to clarify it.}

@@

\response{We have added detail; we believe the presentation of our methods can stand on its own
(although we still relegate minor technical details to the supplementary materials, so as not
to overwhelm the reader). We have also changed presentation more generally so that we don't lead by
saying ``as in Shirreff et al \ldots'', which might make readers think that they need to know Shirreff's
material, although we still try to be clear where we are following Shirreff et al/Champredon et al and
where we are breaking new ground.}

### 2017-02-11 ms/ms.tex
stuff for resubmission

Lastly, we incorporated heterogeneity in sexual activity into the models. Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1 for full model details). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. 
(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix~S2 presents results on the effect of adding heterogeneity to other model variants.)
While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission-virulence_2016}.

@@

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
% 
\begin{thebibliography}{10}

@@

@@

>>>>

Lastly, we incorporated heterogeneity in sexual activity into the models. Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1 for full model details). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. 
(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix~S2 presents results on the effect of adding heterogeneity to other model variants.)
While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission_virulence_2016}.

@@

% \bibliography{virulence}

\begin{thebibliography}{10}

@@

\bibitem{blanquart_transmission_virulence_2016}
Blanquart F, Grabowski MK, Herbeck J, Nalugoda F, Serwadda D, Eller MA, et~al.
\newblock A transmission-virulence evolutionary trade-off explains attenuation
  of {HIV}-1 in {Uganda}.
\newblock eLife. 2016;5:e20492.
\newblock doi:{10.7554/eLife.20492}.

@@

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
%
